A Container-Based Cloud-Native Architecture for the Reproducible Execution of Multi-population Optimization Algorithms

### 2020-10-19 draft.tex
Finished proof reading

A multi-population based algorithm can decrease the execution time
thanks to running the algorithm on several populations at the same
time. But, if the populations are heterogeneous 
it might enhance evolutionary search and avoid unneeded
evaluations to a greater extent than homogeneous systems do; heterogeneous settings, if
done right, increase the diversity of the whole population\cite{araujo2008multikulti}.

But this is a rule of thumb, and it will depend on the degree of heterogeneity,
as well as on the algorithm itself. Some level of heterogeneity can be
implemented by just changing the configuration parameters of each population,
as in the previous experiment. But in this case, we are interested in
answering the second research question, and we want to validate if by using
heterogeneous populations and mixing search strategies,  we can increase the performance of the
search.

Therefore, in this experiment we compare a multi-population with GA and PSO only populations,
versus an ensemble of GA and PSO algorithms. We have tested on the first five functions of the
BBOB testbed; we have used ten populations and eight workers for the experiment and the
same parameters as before.

In order to run this experiment, we just needed to edit a configuration file and
start it up. The plots presented in the results are generated by the BBOB
framework from the JSON log file for each run. There was no additional setup
needed, proving how this framework is able to not only offer a reproducible way
to run experiments, but also how easily it interfaces with other common tools in
the EC ecosystem.

@@

The results obtained by the experiments show (see Figure \ref{fig:bbob}) how
the runtime scales with dimension to reach certain target values $\Delta f$;
The color of the lines indicates the average runtime, for each target reached
at least once. A red circle without a number on top indicates the algorithm
has reached the $10^{-8}$ target on all instances for that dimension. If there
is that number, it indicates how many times the algorithm has reached
the target. Fixed values of targets $\Delta f = 10^{k}$ with $k$ colors are
given in the legend of $f_1$ in the top row, $k$ values are $[-8,-5,-3,-2,-1,0,1]$. Results
from each configuration are presented in each column, GA, PSO, and GA\&PSO
populations.

We can see that the GA algorithm (on the leftmost column) has a hard time reaching even the
$10^{-5}$ target in functions ($f_1-f_4$) for all dimensions, and has the worst
performance in 20 and 40 dimensions. Finally, it performs better on $f_5$
reaching the $10^{-8}$ on all dimensions but with a higher runtime on lower
dimensions.

On the other hand, the multi-population PSO (taking the column on the middle) 
has better performance; its results are spread over a smaller range,
reaching all targets for functions $f_1$ and $f_2$ but with a slightly higher
runtime than the multi-strategy configuration. It is not able to reach the
$10^{-8}$ target on the higher dimensions of $f_3$ and $f_4$. Moreover, the PSO
multi-population has the best runtime for the lower dimensions of $f_5$.

Finally, the proposed PSO\&GA hybrid algorithm has reached the $10^{-8}$ target
value on all functions ($f_1-f_5$) and scales well to higher dimensions, even on
$f_3$ and $f_4$ (charts on the lower rows), which are usually considered
difficult functions. This experiment also exemplifies the type of analyses that
we can perform with the results of the EvoSwarm application by reusing the COCO framework.
We also proved that multi-population algorithms can be compared under the same 
conditions, including single and multi-algorithm versions, and results show 
that the later have better performance.

\section{Conclusions} 
\label{conclusions}

In this paper we have presented the EvoSwarm framework for the
reproducible execution of experiments with an event-based,
cloud-native architecture suitable for hosting multi-population,
multi-algorithm optimization methods.  The framework is based on
industry-standard development tools such as Docker and Docker Compose,
making it easy for scientists and practitioners to develop and deploy
multi-algorithm distributed experiments. The model implemented by the
framework is based on the asynchronous exchange of messages between
stateless functions that react to a continuous stream of data;
populations are the messages that flow in this stream. We show the
differences of the message-driven architecture we presented against
four other proposals for multi-population algorithms found in the
literature, and highlight the advantages for conducting asynchronous
multi-population algorithms.

We have proposed a new methodology for the reproducible deployment and execution of
multiple experiments by specifying the infrastructure as part of an experiment
definition in both local and cloud environments. We presented the workflow of an
experiment's entire life-cycle viewed from the perspective of developers and
researchers. 
We have tested the deployment of an experiment on Amazon's Elastic
Container Service, and proved that as a cloud-native, container-based application,
EvoSwarm could be deployed locally or in the cloud by specifying the configuration 
of container images and resources in an orchestration service.

We have performed an empirical evaluation of the same framework to learn if the proposed
solution can efficiently improve the scalability and performance of
population-based optimization algorithms.

@@

A second experiment centered on evaluating the algorithmic benefits
(experiments taking less time because they need fewer evaluations) of
distributed multi-algorithm algorithms such as the ones implemented in
this kind of framework.  We compared the homogeneous and an ensemble
of multi-populations, using Genetic Algorithms (GAs) and Particle
Swarm Optimization (PSO) using the COCO benchmark framework for
visualization and definition of the optimization problems.  We found
that the set-up of these algorithms has a significant influence on
their results since diversity is essential for their performance. The
hybrid method using both algorithms achieved better results and, in
most cases, needed fewer evaluations.

We conclude that superlinear improvements are achievable by selecting an
appropriate configuration consisting of the number of populations, worker
containers, and algorithm combination. Further research should focus on learning
more about how these parameters interact with each other. The BBOB benchmark
results show that the selection and mixing of algorithms can lower the number of
FEs needed to find a solution or find solutions that single algorithms could not
find.

This work also opens new possibilities. From the design point of view, it would
be interesting to use automatic scaling, instead of setting the number of
workers by hand. This would need a certain amount of configuration, as we would
be using container orchestration with auto-scaling capabilities such as
Kubernetes or Docker Swarm. The current orchestration (Amazon ECS) also has
auto-scaling capabilities, but additional configuration is needed; also, we need
to establish the auto-scaling of both containers and Amazon's EC2 instances.
Instead of setting a fixed number of workers, only the maximum amount would need
to be established, which could be done directly or by setting an evaluation
budget.

From the algorithmic point of view, there are many possible lines of
research. Mixing population-based algorithms with other kinds
would be a possibility, but also using different instances of
population based algorithms, such as checking how Estimation of
Distribution Algorithms would work together with evolutionary
algorithms. EvoloPy also includes other population based algorithms,
which could be tested, trying to find out which sets of algorithms are
a better match for each other. All these will be explored as future
lines of work.

>>>>

A multi-population based algorithm can decrease the execution time
thanks to running the algorithm on several populations simultaneously.
However, if the populations are heterogeneous 
it might enhance the evolutionary search and avoid unneeded
evaluations to a greater extent than homogeneous systems do; heterogeneous settings, if
done right, increase the diversity of the whole population\cite{araujo2008multikulti}.

Nevertheless, this is a rule of thumb, and it will depend on the degree of heterogeneity and 
the algorithm itself. Some level of heterogeneity can be 
achieved by just changing the configuration parameters of each population, 
as in the previous experiment.However, in this case, we are interested in answering 
the second research question. We want to validate if we can increase the search's performance 
using heterogeneous populations and mixing search strategies.

Therefore, in this experiment, we compare a multi-population with GA and PSO only populations,
versus an ensemble of GA and PSO algorithms. We have tested on the first five functions of the
BBOB testbed; we have used ten populations and eight workers for the experiment and the
same parameters as before.

To run this experiment, we just needed to edit a configuration file and
start it up. The plots presented in the results are generated by the BBOB
framework from the JSON log file for each run. There was no additional setup
needed, proving how this framework can offer a reproducible way
to run experiments and how easily it interfaces with other common tools in
the EC ecosystem.

@@

The results obtained by the experiments show (see Figure \ref{fig:bbob}) how
the runtime scales with dimension to reach certain target values $\Delta f$;
The color of the lines indicates the average runtime, for each target, reached
at least once. A red circle without a number on top indicates the algorithm 
reached the $10^{-8}$ target on all instances for that dimension. If there
is a number, it indicates how many times the algorithm has reached
the target. Fixed values of targets $\Delta f = 10^{k}$ with $k$ colors are
given in the $f_1$ legend in the top row, $k$ values are $[-8,-5,-3,-2,-1,0,1]$. Results
from each configuration are presented in each column, GA, PSO, and GA\&PSO
populations.

We can see that the GA algorithm (on the leftmost column) has a hard time reaching even the
$10^{-5}$ target in functions ($f_1-f_4$) for all dimensions and has the worst
performance in 20 and 40 dimensions. Finally, it performs better on $f_5$
reaching the $10^{-8}$ on all dimensions but with a higher runtime on lower
dimensions.

On the other hand, the multi-population PSO (taking the column in the middle) 
has better performance; its results are spread over a smaller range,
reaching all targets for functions $f_1$ and $f_2$ but with a slightly higher
runtime than the multi-strategy configuration. It is not able to reach the
$10^{-8}$ target on higher dimensions of $f_3$ and $f_4$. Moreover, the PSO
multi-population has the best runtime for the lower dimensions of $f_5$.

Finally, the proposed PSO\&GA hybrid algorithm has reached the $10^{-8}$ target
value on all functions ($f_1-f_5$). It scales nicely to higher dimensions, even on
$f_3$ and $f_4$ (charts on the lower rows), usually considered
difficult functions. This experiment also exemplifies the type of analyses that
we can perform with the EvoSwarm application results by reusing the COCO framework.
We also proved that multi-population algorithms could be compared under the same 
conditions, including single and multi-algorithm versions, and results show 
that the latter has the best performance.

\section{Conclusions} 
\label{conclusions}
This paper has presented the EvoSwarm framework for the
reproducible execution of experiments with an event-based,
cloud-native architecture suitable for hosting multi-population,
multi-algorithm optimization methods.  The framework is based on
industry-standard development tools such as Docker and Docker Compose,
making it easy for scientists and practitioners to develop and deploy
multi-algorithm distributed experiments. The model implemented by the
framework is based on the asynchronous exchange of messages between
stateless functions that react to a continuous stream of data;
populations are the messages that flow in this stream. We show the
differences of the message-driven architecture we presented against
four other proposals for multi-population algorithms found in the
literature and highlight the advantages of conducting asynchronous
multi-population algorithms.

We have proposed a new methodology for the reproducible deployment and execution of
multiple experiments by specifying the infrastructure as part of an experiment
definition in both local and cloud environments. We presented the workflow of an
experiment's entire life-cycle viewed from the perspective of developers and
researchers. 
We have tested the deployment of an experiment on Amazon's Elastic
Container Service and proved that as a cloud-native, container-based application,
EvoSwarm could be deployed locally or in the cloud by specifying the configuration 
of container images and resources in an orchestration service.

We have performed an empirical evaluation of the same framework to learn if the proposed
solution can efficiently improve population-based optimization algorithms' 
scalability and performance.

@@

A second experiment centered on evaluating the algorithmic benefits
(experiments are taking less time because they need fewer evaluations) of
distributed multi-algorithm algorithms such as the ones implemented in
this kind of framework.  We compared the homogeneous and an ensemble
of multi-populations, using Genetic Algorithms (GAs) and Particle
Swarm Optimization (PSO), using the COCO benchmark framework to visualize and define 
optimization problems.  We found
that these algorithms' set-up has a significant influence on
their results since diversity is essential for their performance. The
hybrid method using both algorithms achieved better results and, in
most cases, needed fewer evaluations.

We conclude that superlinear improvements are achievable by selecting an
appropriate configuration consisting of the number of populations, worker
containers, and algorithm combinations. Further research should focus on learning
more about how these parameters interact with each other. The BBOB benchmark
results show that the selection and mixing of algorithms can lower the number of
FEs needed to find a solution or find solutions that single algorithms could not
find.

This work also opens new possibilities. From the design point of view, it would
be interesting to use automatic scaling, instead of setting the number of
workers by hand. This would need a certain amount of configuration, as we would
be using container orchestration with auto-scaling capabilities such as
Kubernetes or Docker Swarm. The current orchestration (Amazon ECS) also has
auto-scaling capabilities, but additional configuration is needed; we also need
to establish the auto-scaling of both containers and Amazon's EC2 instances.
Instead of setting a fixed number of workers, only the maximum amount would need
to be established, which could be done directly or by setting an evaluation
budget.

From the algorithmic point of view, there are many possible lines of
research. Mixing population-based algorithms with other kinds
would be a possibility and using different instances of
population-based algorithms, such as checking how Estimation of
Distribution Algorithms would work together with evolutionary
algorithms. EvoloPy also includes other population-based algorithms,
which could be tested, trying to determine which sets of algorithms are
a better match for each other. We will explore all this as future lines of work.
